Optimal synaptic strategies for different timescales of memory

### 2016-03-10 Paper/SynapsePaper.tex
some ref fixing

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage{graphicx}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
%\usepackage{pnastwoF}
\usepackage{amssymb,amsfonts}
%\usepackage[centertags]{amsmath}
%\usepackage{hyperref}

%% OPTIONAL MACRO DEFINITIONS
\input{sl_definitions.tex}
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

@@

\author{Subhaneil Lahiri\affil{1}{Department of Applied Physics, Stanford University, Stanford CA},
\and
Surya Ganguli\affil{1}{}}

@@

\abbreviations{LTP, long term potentiation; 
LTD, long term depression}

@@

A simple average memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ (or $\eq\M\dep$) in those synapses that experience a candidate potentiation (or depression) event.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{equation}\label{eq:evol}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{equation}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in the supplementary material.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  }\label{fig:models}
\end{figure}

In this work, rather than analyzing specific models,  we take a different approach, in order to obtain a more general theory.
We consider the \emph{entire} space of these models and find upper bounds on the memory capacity of \emph{any} of them.
The space of models with a fixed number of internal states $M$ is parameterized by the pair of $M \times M$ discrete time stochastic
transition matrices $\M\pot$ and $\M\dep$, in addition to $f\potdep$.  The parameters must satisfy the following constraints:
%
\begin{equation}\label{eq:constr}
\begin{aligned}
  \M\potdep_{ij} &\in [0,1], &\quad
  f\potdep &\in [0,1], &\quad
  \eq\frg &= 0, &\quad
  \w_i &= \pm 1, \\
  \sum_j \M\potdep_{ij} &= 1, &
  f\pot + f\dep &= 1, &
  \sum_i \eq_i &= 1.
\end{aligned}
\end{equation}
%
The upper bounds on $\M\potdep_{ij}$ and $f\potdep$ follow automatically from the other constraints.

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
  \frg = \sum_a -q_a \uv^a \vv^a,
  \quad
  \vv^a \uv^b = \delta_{ab},
  \quad
  \frg \uv^a = -q_a \uv^a,
  \quad
  \vv^a \frg = -q_a \vv^a.
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved in \sref{sec:env}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.12\linewidth]{max_init_pot.svg}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.12\linewidth]{max_init_dep.svg}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.05\linewidth]{binary_det.svg}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  }\label{fig:max_init}
\end{figure}

@@

First passage time theory introduced in the \hyperref[sec:order]{previous section} becomes useful because the area has a simple expression in terms of quantities introduced in \eqref{eq:kemenypm} (see supplementary material):
%
\begin{equation}\label{eq:areaEta}
\begin{aligned}
  A &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \prn{\M\pot_{ij}-\M\dep_{ij}} \prn{\eta^+_i-\eta^+_j}\\
    &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \prn{\M\pot_{ij}-\M\dep_{ij}} \prn{\eta^-_j-\eta^-_i}.
\end{aligned}
\end{equation}
%

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.28\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  }\label{fig:perts}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \Fref{fig:env}(\ref{fig:sticky}). %\Fref{fig:max_area}.
This makes it very difficult to exit these states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

\bibliographystyle{pnas2011}
\bibliography{maths,neuro}

>>>>

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage{graphicx}
\usepackage{ifpdf}
%\usepackage[hypcap]{caption}%links to figures
\usepackage{graphicx}
%\usepackage{graphics} for finding documentation only
%
\usepackage{epstopdf}
\epstopdfsetup{update,suffix=-generated}
%
%\newcommand{\inkscapedir}{C:/Program Files/Inkscape/}
%\IfFileExists{"C:/Program Files (x86)/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Program Files (x86)/Inkscape/}}{}
\IfFileExists{"C:/Program Files/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Program Files/Inkscape/}}{}
\IfFileExists{"/Applications/Inkscape.app/Contents/Resources/bin/inkscape"}{\newcommand{\inkscapedir}{/Applications/Inkscape.app/Contents/Resources/bin/}}{}
%
\ifpdf
 \epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{%
 "\inkscapedir inkscape" -f #1 -C -A \OutputFile
 }
\else
 \DeclareGraphicsRule{.png}{eps}{.bb}{}
\fi
%
\usepackage{adjustbox}
\newcommand{\aligntop}[1]{\adjustbox{valign=t}{#1}}
\newcommand{\alignmid}[1]{\adjustbox{valign=c}{#1}}
%
%\usepackage{paralist}
%\newenvironment{myenuma}{\begin{inparaenum}[(a)]}{\end{inparaenum}}
%\newenvironment{myenumi}{\begin{inparaenum}[i)]}{\end{inparaenum}}
%\newenvironment{myenumA}{\begin{inparaenum}[\bfseries A.]}{\end{inparaenum}}
%\newenvironment{myenumI}{\begin{inparaenum}[\bfseries I.]}{\end{inparaenum}}
%\newenvironment{myenumr}{\begin{inparaenum}[1.]}{\end{inparaenum}}
%%for nested lists, \ref only produces innermost item:  (hacky)
%\makeatletter % for internal macros with @
%\renewcommand\p@enumii{}
%\makeatother
%
\usepackage[inline]{enumitem}
\newlist{myenuma}{enumerate*}{4}
\setlist[myenuma]{label=(\alph*),ref=\alph*}
%\makeatletter
%\DeclareRobustCommand{\ilabel}[1]{%
%   \edef\@currentlabel{\themyenumai}%
%   \label{#1}}
%\makeatother
\graphicspath{{"Figs/"}}
%\usepackage{pnastwoF}
%\usepackage{amssymb,amsfonts}
%\usepackage[centertags]{amsmath}
%\usepackage{hyperref}

%% OPTIONAL MACRO DEFINITIONS
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
% indices
%
% brackets etc.
\newcommand{\prn}[1]{\left ( #1 \right )}
\newcommand{\brc}[1]{\left\{ #1 \right\}}
\newcommand{\brk}[1]{\left [ #1 \right ]}
\newcommand{\abs}[1]{\left\lvert #1 \right\rvert}
\newcommand{\nrm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\avc}[2]{\left\langle #1 \,\middle | \, #2 \right\rangle}
%\newenvironment{cases}{\left\{\begin{aligned}}{\end{aligned}\right.}
%
% Sets
\newcommand{\set}[2]{\left\{ #1 \,\middle |\, #2 \right\}}
\newcommand{\cond}[2]{\!\left( #1 \,\middle | \, #2 \right)}
\newcommand{\condb}[2]{\!\left[ #1 \,\middle | \, #2 \right]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][\rule{0mm}{0mm}]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][\rule{0mm}{0mm}]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][\rule{0mm}{0mm}]{\left (\frac{\partial #2}{\partial {#3}}\right )_{\!\!#1}}
\newcommand{\intd}[2][\rule{0mm}{0mm}]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
% Referencing sections, figures, etc
\newcommand{\sref}[1]{\S\ref{#1}}
\newcommand{\fref}[1]{fig.\ref{#1}}
\newcommand{\Fref}[1]{Fig.~\ref{#1}}
\newcommand{\tref}[1]{tab.\ref{#1}}
\newcommand{\Tref}[1]{Tab.\ref{#1}}
%
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

@@

\author{Subhaneil Lahiri\affil{1}{Department of Applied Physics, Stanford University, Stanford CA}
\and
Surya Ganguli\affil{1}{}}

@@

\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation; 
LTD, long term depression}

@@

A simple average memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ (or $\eq\M\dep$) in those synapses that experience a candidate potentiation (or depression) event.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{multline}\label{eq:evol}
  \diff{\prob(t)}{t} = r\prob(t)\frg,
  \\ \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{multline}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}(2 f\pot f\dep) \eq \brk{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in the supplementary material.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  \label{fig:models}}
\end{figure}

In this work, rather than analyzing specific models,  we take a different approach, in order to obtain a more general theory.
We consider the \emph{entire} space of these models and find upper bounds on the memory capacity of \emph{any} of them.
The space of models with a fixed number of internal states $M$ is parameterized by the pair of $M \times M$ discrete time stochastic
transition matrices $\M\pot$ and $\M\dep$, in addition to $f\potdep$.  The parameters must satisfy the following constraints:
%
\begin{equation}\label{eq:constr}
\begin{gathered}
\begin{aligned}
  \M\potdep_{ij} &\in [0,1], &\qquad
  \sum_j \M\potdep_{ij} &= 1, \\
  f\potdep &\in [0,1],&
  f\pot + f\dep &= 1, 
\end{aligned}
\\
  \eq\frg = 0, \qquad
  \sum_i \eq_i = 1, \qquad
  \w_i = \pm 1.
\end{gathered}
\end{equation}
%
The upper bounds on $\M\potdep_{ij}$ and $f\potdep$ follow automatically from the other constraints.

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[sec:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.24\linewidth]{max_init_pot.svg}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.24\linewidth]{max_init_dep.svg}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.1\linewidth]{binary_det.svg}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  \label{fig:max_init}}
\end{figure}

@@

First passage time theory introduced in the \hyperref[sec:order]{previous section} becomes useful because the area has a simple expression in terms of quantities introduced in \eqref{eq:kemenypm} (see supplementary material):
%
\begin{equation}\label{eq:areaEta}
\begin{aligned}
  A &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^+_i-\eta^+_j)\\
    &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^-_j-\eta^-_i).
\end{aligned}
\end{equation}
%

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.28\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  \label{fig:perts}}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
\begin{aligned}
  A &\leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k \\
    &= \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{aligned}
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \Fref{fig:env}(\ref{fig:sticky}). %\Fref{fig:max_area}.
This makes it very difficult to exit these states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

\bibliographystyle{pnas2011}
\bibliography{maths,neuro,markov}
